Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 RNARedPrint.tex
SOme edits

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasiblity of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\parHead{Specificities and similarities of multi-target design.}
The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
%
Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
%
Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
search in \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the \emph{uniform} distribution for multiple
target structures. \Software{RNAdesign} adopts a graph coloring perspective,
coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors. Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform \Def{stochastic backtracking}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. While, due to its decomposition strategy, 
\Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility. Both aspects call for a more systematic redesign of multi-target sampling, which builds on established concepts in computer science. In turn, this promises a fundamentally broader range of applications for extensions of the sampling method.

@@

\begin{figure*}[t]
\begin{center}
    \includegraphics[width=.8\textwidth]{Figs/Workflow}
\end{center}
\caption{General outline of \ourprog{}. From a set of target secondary structures (i), base pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi). \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures. Our computations based on a more complex energy model, which considers energy contributions of base pair stacks, require additional dependencies.}
  }
\label{fig:workflow}
\end{figure*}

\revised{As central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features. For the sake of clarity, we introduce this method specialized to the sampling of RNA sequences that target the energies of multiple structures and the \GCb\%.}

\revised{Figure~\ref{fig:workflow} summarizes our generic framework, which enables this targeted sequence generation based on multi-dimensional Boltzmann sampling. \emph{Algorithmically}, we originally contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
.
\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies induced by the target structures) to guide the efficient processing by DP algorithms.}
%
\revised{Building on this principle, the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}. Thus, it is essential for the applicability of our approach that---by appropriate design choices---we can keep this parameter low for typical instances. For} any fixed value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are \emph{fixed-parameter tractable (FPT)}.

\revised{Remarkably, we could show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard. The practical relevance of this theoretical result is that it rules out substantially better sampling techniques, such that there will always---even using improved sampling methods---remain an upper limit on the---in practice---tractable number and heterogeneity of structures, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.

% Subsection~\ref{sec:complexity}
%
Due to the generality of our method, we can moreover strongly limit \revised{the treewidth} in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating sequences in a specialized weighted constraint
network, we support---in principle---arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

We observe that
targeting realistic RNA energies in the
Turner RNA energy model
works well by performing sampling based on \revised{a simplified RNA energy model, which induces much lower treewidth than the Turner model}. This result is \revised{essential for the applicability of our method, since it allows to combine high efficiency (by keeping the treewidth low) with sufficient accuracy to precisely target realistic Turner energies.}

\revised{Eventually our proof-of-concept results 
 on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
  suggest that our sampling strategy well supports designing biologically relevant
RNAs for multiple targets.}

\section*{Methods}
\revised{The main computational problem addressed in this work is the positive design of RNA sequences for multiple target structures; more specifically, the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
Here, these desired sequence properties are modeled as constraints on the values of \Def{features}, which are functions of the sequence that are expressed as sums over real-valued \Def{contributions}. Each contribution depends on the nucleotides at---typically few---specific sequence positions.}

\revised{To generate diverse design candidates, we randomly generate sequences from a Boltzmann distribution. The probability of a sequence then depends on its features (e.g. the energies of the target structures), and the weight of each feature (which influences its distribution). Sampling from the (multi-feature) Boltzmann distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight. 
On this basis, we can finely calibrate the weights, to maximize the probability that sampled sequences meet the desired target values for each feature. Together with a final rejection step this results in an effective procedure for generating highly specific sequences.}

\subsection*{Problem statement}
\label{sec:problem-statement}

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as: 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

@@

The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}

\revised{
  To start with a simple example of evaluating the energy of sequences by features, let us explain how they are used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$. Consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that 
	$$f^{{\sf BP}}_{i,j}(\left\{\substack{i\mapsto a\\j\mapsto b}\right\})=\begin{cases}0&\text{ if }\{a,b\}\in\mathcal{B} \\+\infty&\text{ otherwise}.\end{cases}$$ 
        The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created. For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Eq.~\ref{eq:mainproblem} (when restricted to $F^{\tt BP}) $simply counts the number of valid sequences.

        Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, base pair stacks, loops\ldots) under a certain nucleotide assignment. Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$. For instance, the contribution of a \Def{base pair stack}, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.
        We refer to energy models that consider the contributions of all base pair stacks (and thus introduce the corresponding dependencies) collectively as the \Def{stacking energy model}.
}

@@

\subsection*{Dependency (hyper)graph, tree decomposition and treewidth}

%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined way, one must consider dependencies induced by the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$}

%
%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
%

\revised{
In the simplest case, this set captures the requirement of canonical base pairing for each structure.
To express this, let us define the \Def{base pair dependency graph} $G_{\R}$ as the graph with nodes $\{1,\dots,n\}$
and edges $\bigcup_{\ell\in[1,k]} R_\ell$. %Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in some structure, inducing a direct dependency between the chosen nucleotide assignments for the two positions.

Since $\F$ defines potentially more complex dependencies, which can
relate more than two positions, in general its dependencies cannot be 
represented by a graph. Instead, this requires a structure known as
\Def{hypergraph}, which consists of vertices (here, the sequence
positions) connected by \Def{hyperedges}, which are arbitrary sets of
vertices. In this way, hypergraphs generalize undirected graphs, where
each edge is a set of exactly two vertices.
%
The \Def{dependency (hyper)graph induced by $\F$} is then defined as the hypergraph $G_\F=(V, H)$ on sequence positions $V=\{1,\dots,n\}$
by interpreting the dependencies as hyperedges, i.e. $H =
\{\dep(f)\mid f\in \F\})$.

Let us finally define the \Def{tree decomposition} of the graph $G_\F$, a fundamental ingredient of our algorithms, which also determines their efficiency (most importantly, via its property called treewidth).
}
\begin{definition}[Tree decomposition and treewidth]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a (hyper)graph with nodes in $X$ and (hyper)edges in $E$. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest and, for
  each $v\in T$, $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}

\smallskip
\noindent
The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
$\min_{u\in T} |\chi(u)| - 1.$ 
%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
\end{definition}
%
\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependend subproblems to ensure consistency in a recursive computation. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

\begin{figure}
{\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model (where the four positions of each base pair stack depend on each other). Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in base pair stacks (colored). A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each base pair stack is represented in at least one of its node, so that features can be correctly evaluated. The treewidth of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
\end{figure}

@@

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp.\ Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space; Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp.\ Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.

>>>>

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility. \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} For this task, we present the fixed-parameter tractable (FPT) approach \ourprog{}.
  % 
  \revised{Based on} constraint networks, \ourprog{} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  %Securing the theoretical foundations of our method
  Moreover, we establish the $\#${\sf P}-hardness of \revised{counting to theoretically justify the development of an FPT approach}. 
%
  Finally, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  common design benchmark.
  We demonstrate significant improvements over the previously best multi-target sampling strategy, due to generating seed sequences with \ourprog{}.\\
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\parHead{Specificities and similarities of multi-target design.}
The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
%
Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
%
Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
search in \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the \emph{uniform} distribution for multiple
target structures. \Software{RNAdesign} adopts a graph coloring perspective,
coloring the sequence positions with nucleotides, such that the ends of each base pair receive compatible colors. Initially the method decomposes the graph hierarchically, using ear decomposition, and then \Def{precomputes} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform \Def{stochastic backtracking}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. While, due to its decomposition strategy, 
\Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility. Both aspects call for a more systematic redesign of multi-target sampling, which builds on established concepts in computer science. In turn, this promises a fundamentally broader range of applications for extensions of the sampling method.

@@

\begin{figure*}[t]
\begin{center}
    \includegraphics[width=.8\textwidth]{Figs/Workflow}
\end{center}
\caption{General outline of \ourprog{}. From a set of target secondary structures (i), base pairs are merged (ii) into a (base pair) dependency graph (iii) and transformed into a tree decomposition (iv). The tree is then used to compute the partition function, followed by a Boltzmann sampling of valid sequences (v). An adaptive scheme learns weights to achieve targeted energies and \GCb\%, leading to the production of suitable designs (vi). \revised{Note that for simplicity, we assume in this figure that only dependencies between the ends of base pairs are considered to evaluate the energies of structures. Our computations based on a more complex energy model (stacking model) require additional dependencies.}
  }
\label{fig:workflow}
\end{figure*}

\revised{As our central contribution, we provide a systematic and flexibly extensible technique for sampling that targets multiple versatile features. For the sake of clarity, we introduce this method specialized to the sampling of sequences that target the energies of multiple structures and the \GCb\%.}

\revised{Fig.~\ref{fig:workflow} summarizes our generic framework, which enables} efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures. \revised{\emph{Algorithmically}, we contribute} dynamic programming (DP) algorithms, based on the concept of \Def{tree decomposition}, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
.
\revised{Generally, tree decompositions are data structures that capture the specific dependencies of a problem instance (here, the dependencies induced by the target structures) to guide the efficient processing by DP algorithms.}
%
\revised{Building on this principle, }the complexities of our algorithms depend exponentially on a specific property of the tree decomposition, called the \Def{treewidth}, which remains low for typical instances. However, for any given value of the treewidth, the complexity scales only linearly with the size of designed sequences and the number of targeted structures, \emph{i.e.} our algorithms are fixed-parameter tractable (FPT).

\revised{Remarkably, we show that it is not possible to find a better, efficient method for sampling (unless ${\sf P}={\sf NP}$), since the underlying counting problem is \#{\sf P}-hard. The practical relevance of this result is that it rules out substantially better sampling techniques, such that there will always remain---in practice---an upper limit on the tractable number of structures and their heterogeneity, the complexity of the directly treatable energy model, and the number and complexity of additional constraints that could be considered in future sampling-based applications.} Technically, this result relies on a surprising bijection between valid sequences and independent sets of a bipartite graph, the latter being the object of recent breakthroughs in approximate counting complexity~\citep{Bulatov2013,Cai2016}.

% Subsection~\ref{sec:complexity}
%
Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a specialized weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

We observe that
targeting realistic RNA energies in the
Turner RNA energy model
works well by performing sampling based on \revised{a simplified RNA energy model, which induces lower treewidth}. This result is highly relevant in practice, since it allows to combine high efficiency with high accuracy (and even leaves room for \revised{future extensions supporting additional constraints}).

\revised{Eventually our proof-of-concept results 
 on a comprehensive multi-target RNA design benchmark~\citep{Taneda2015} 
  suggest that our sampling strategy supports designing biologically relevant
RNAs for multiple targets.}

\section*{Methods}
The main computational problem addressed in this work is the generation of sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$, \revised{such that the sequences feature a given $\GCb\%$, and have prescribed energies for a set of target secondary structures.
Here, such criteria are modeled} as constraints on the value of \Def{features}, functions of the sequence that are expressed as sums over real-valued contributions. Such contributions typically depend on the nucleotides assigned to a subset of sequence positions.

To achieve maximal diversity within candidate designs, we abstract the problem as a random generation of sequences within a Boltzmann distribution. The probability of a sequence then depends on its associated features, and of weights that are associated features. Those influence the expected feature values of generated sequences, allowing us to shift the sample distributions towards specific targets. Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight. A final rejection step, made efficient by the fine calibration of weights, allows to produce sequences satisfying the input set of constraints.

\subsection*{Problem statement}
\label{sec:problem-statement}
%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and $k$ features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$, our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as: 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences:
%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

@@

The $\GCb\%$ can be trivially expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
\revised{By summing $f^{\GCb}_i$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}

\revised{
	Such features can also be used to count the number of \Def{valid sequences}, \emph{i.e.} sequences inducing only base pairs in $\mathcal{B}:=\{\{\Ab,\Ub\}, \{\Gb,\Cb\}, \{\Gb,\Ub\}\}$. Indeed, consider a feature $F^{\tt BP}$ composed of  contributions $f^{\tt BP}_{i,j}$, for each base pair $(i,j)$ occurring in some structure, such that 
	$$f^{{\sf BP}}_{i,j}(\left\{\substack{i\mapsto a\\j\mapsto b}\right\})=\begin{cases}0&\text{ if }\{a,b\}\in\mathcal{B} \\+\infty&\text{ otherwise}.\end{cases}$$ 
	The value of $F^{\tt BP}$ is $0$ for any valid sequence, and $+\infty$ as soon as some non canonical base pair is created. For any associated weight $\pi_{{\sf BP}}>1$, the contribution of a valid sequence is $\pi_{{\sf BP}}^0 = 1$, and the contribution of an invalid sequence is $\pi_{{\sf BP}}^{+\infty} = 0$, so that Equation~\ref{eq:mainproblem} simply counts the number of valid sequences when restricted to $F^{\tt BP}$.

	Energy models for structure prediction vary considerably, yet can always be expressed as sums over contributions associated with local structural motifs (base pairs, stacking pairs, loops\ldots) under a certain nucleotide assignment. Energy models can thus be captured generically by introducing, for each motif $m$ occurring in a target structure, a contribution $f_m$, taking a specific value for each assignment of nucleotides to its positions $\dep(f_m)$. For instance, the contribution of a stacking pair, consisting of two pairs $(i,j)$ and $(i+1,j-1)$, can be captured by the introduction of a function $f^{\textsf{Stack}}_{\substack{i,j}}$ such that $\dep(f^{\textsf{Stack}}_{\substack{i,j}}) = \{i,i+1,j-1,j\}$.

}

@@

\subsection*{Dependency (hyper)graph, tree decomposition and width}
%The complexity of the central problem depends fundamentally on the dependency structure that is imposed by the entire set of contributions.

@@

\revised{In order to compute the partition function of Eq.~\ref{eq:mainproblem}, and thus sample in a well-defined distribution, one needs to abstract dependencies beyond pairs of positions. To that purpose, we consider the notion of \Def{hypergraph}, which consists of vertices (here, the sequence positions) connected by \Def{hyperedges} defined as sets of vertices. Undirected graphs can be modeled as hypergraphs whose hyperedges are pairs of vertices, but hypergraphs also allow to express complex dependencies.

Here, we consider the complete set of contributions $$\F := \bigcup_{\ell}\{f\mid f\text{ contribution of } F_\ell\}.$$
%
%Let us denote the \emph{dependencies of a contribution $f$} as \dep{f}.
%
The set $\F$ of contributions induces a
\Def{dependency (hyper)graph} $G_\F$ on sequence positions by interpreting the dependencies as hyperedges, leading to an hypergraph 
$G_\F=(V, H)$ having vertices $V=\{1,\dots,n\}$ and hyperedges $H = \{\dep(f)\mid f\in \F\})$.

Irrespectively of the chosen energy model, to capture the requirements of canonical base pairing for each structure,
we also define the \Def{base pair dependency graph} $G_{\R} = (V,E)$ having vertices $V=\{1,\dots,n\}$ 
and edges $E=\bigcup_{\ell\in[1,k]} R_\ell$. Any edge between two nodes $i$ and $j$ materializes the existence of a base pair between positions $i$ and $j$ in at least one of the input structures, and induce a direct dependency between the nucleotide assignments for those positions.
}

Let us now remind the \Def{tree decomposition} of an (hyper)graph $G_\F$, a fundamental ingredient of our algorithms.

\begin{definition}[Tree decomposition]
  \label{def:treedecomp}
  Let $G=(X, E)$ be a (hyper)graph with nodes in $X$ and (hyper)edges in $E$. A \Def{tree decomposition} of $G$ is
  a pair $(T,\chi)$, where $T$ is an unrooted tree/forest and, for
  each $v\in T$, $\chi(v)\subseteq X$ is a set of vertices assigned to
  the node $v\in T$, such that
\begin{enumerate}
\item each $x\in X$ occurs in at least one $\chi(v)$;
\item for all $x\in X$, $\{ v \mid x \in \chi(v) \}$ induces a connected subtree of $T$;
\item for all $e\in E$, there is a node $v\in T$, such that $e\subseteq\chi(v)$.
\end{enumerate}
\end{definition}
\revised{Intuitively, a tree decomposition for an (hyper)graph $G$ is a tree structure which features all the vertices of $G$, captures all the (hyper)edges of $G$, and helps ensure some notion of consistency by allowing the propagation of values. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}
\begin{definition}[Tree width]
The \Def{width} of a tree decomposition $(T,\chi)$ is 
$$\min_{u\in T} |\chi(u)| - 1.$$ The \Def{treewidth}
of an (hyper)graph $G$ is the smallest width of a tree decomposition of $G$.
\end{definition}
%Recall the essential role of the treewidth as fixed parameter in our FPT appraoch.

\begin{figure}
{\centering \includegraphics[width=.9\linewidth]{Figs/example-treedec-stacking}\\}
\caption{\revised{Toy example of a tree decomposition associated with two target structures in the stacking energy model. Two target secondary structures ({\bfseries\sffamily A}) are merged into a joint hypergraph ({\bfseries\sffamily B}), whose hyperedges correspond to the quadruplets of positions involved in stacking pairs (colored). A valid tree decomposition ({\bfseries\sffamily C}) for the hypergraph ensures, among other properties, that each base pair and each stacking pair/hyperedge is represented in at least one of its node, so that features can be adequately evaluated. The width of this tree decomposition is 3, a provably optimal value for this input hypergraph.}\label{fig:treedec}}
\end{figure}

@@

Algorithm~\ref{alg:pf} computes the partition function by passing
messages along the directed edges $u\to v$ (which point from 
child $u$ to its parent $v$). Each message $m$ has the form of a contribution, i.e. it takes a partial sequence, depends
on the positions $\dep(m)\subseteq \X$, and yields a partition function
in $\real$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function (i.e.~due to proper encoding the partition function of our design problem) through
\begin{math}
  \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

%The complexity of the proposed sampling algorithm depends on the treewidth of the dependency graph $G_\F$.
% The treewidth $w$ of a cluster tree $(T,\chi,\phi)$, $T=(V,E)$, is $\max_{v\in V} |\chi(v)| - 1$, i.e.~the maximum number of positions in any of its clusters minus 1.
Let $s$ denote the maximum size of any separator set $\separator{u}{v}$ and $D$ denote the maximum size of
$\difference{u}{v}$ over $(u,v)\in E$.  In the absence
of specific optimizations, running Alg.~\ref{alg:pf} requires
$\mathcal{O}((|\F|+|V|)\cdot 4^{w+1})$ time and
$\mathcal{O}(|V|\cdot4^s)$ space
(Supp. Mat.~\ref{appsec:algcomplexity}); Alg.~\ref{alg:sampling} would
require $\mathcal{O}((|\F|+|V|)\cdot 4^D)$ per sample on arbitrary
tree decompositions
(Supp. Mat.~\ref{appsec:algcomplexity}). W.l.o.g.\ we assume that
$D=1$; note that tree decompositions can generally be transformed,
such that $\difference{u}{v}\leq 1$.
%
Moreover, the size of $\F$ is linearly bounded: for $k$
input structures for sequences of length $n$, the energy function is
expressed by $\mathcal{O}(n\,k)$ functions. Finally, the number of cluster
tree nodes is in $O(n)$, such that $|\F|+|V| \in \mathcal{O}(n\,k)$.
